Cartograms with Dynamic Features

### 2022-03-11 paper/section/data.tex

Clinical Commissioning Groups (CCGs) are the main geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, because there is still no public EHR data published based on the latest CCG re-organization that took place in 2021.

@@

We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for River Thames, River Trent and River Great Ouse in England. These rivers were chosen as they pass through regions with dense population, and provide informative geographical and topological cues.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections and scales, and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file to reduce the file size and improve the performance of our implementation.

@@

>>>>

Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, because there is still no public EHR data published based on the latest CCG re-organization that took place in 2021.

@@

We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for River Thames, River Trent and River Great Ouse in England. These rivers were chosen as they are well-known rivers and pass through regions with dense population, and provide informative geographical and topological cues.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections and scales, and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file. TopoJSON eliminates redundant coordinates in the data, this improves the rendering speed of our implementation.

@@

Each CCG has a unique ONS code, which is used to link the CCG shapefile with the statistical data.

### 2022-03-11 paper/section/introduction.tex

Cartograms are representations of geographical space generated by a value-by-area mapping technique that combines statistical and geographical information \cite{dent2009Cartography}. \new{Various styles of cartograms have been proposed and implemented, covering topics on urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision making \cite{gao2020Visualising, sack2021Visualizing}.}

In this paper, we follow the categorization of cartograms by \citea{nusrat2016State}: contiguous, non-contiguous, rectangular, and Dorling. Among these four types, trade-offs between accuracies are made (See \tableref{table:accuracy}). Building on Demers cartograms \cite{ian2002Cartogram}, we develop dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We incorporate a cartographic layout algorithm that includes these topological features to update the layout of the nodes representing geographical regions. To minimize geographical errors, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm a real world Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm.

@@

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability.
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers.
\end{itemize}

>>>>

Cartograms are representations of hybrid geographical and data space based on a value-by-area mapping technique that combines statistical and geographical information \cite{dent2009Cartography}. Various styles of cartograms have been proposed and implemented, covering applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision making \cite{gao2020Visualising, sack2021Visualizing}.

In this paper, we follow the categorization of cartograms by \citea{nusrat2016State}: contiguous, non-contiguous, rectangular, and Dorling. Among these four types, a trade-off between accuracy is made (See \tableref{table:accuracy}). For this project we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We implement a cartographic layout algorithm that includes these topological features into the layout of the nodes representing geographical regions. To minimize geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm a real world case-study using an Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm.

@@

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability.
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers.
    \item A user study evaluation of the technique with an application to EHRs
\end{itemize}

### 2022-03-11 paper/section/related.tex
draft: update figures and algorithms

While we focus on rectangular and Dorling cartograms in this paper, we provide brief definitions of contiguous and non-contiguous cartograms for interested readers: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent neighbors but have distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity with neighbors to enable growing or shrinking in size while maintaining their original shape \cite{dent2009Cartography}.

@@

{
\renewcommand{\arraystretch}{1.5}
\begin{table}[!b]
	\centering
	\resizebox{\columnwidth}{!}{
		\begin{tabulary}{\columnwidth}{|*{5}{l|}}
			\hhline{~|*{3}{-}|~}
			\multicolumn{1}{c|}{} & \multicolumn{3}{c|}{\cellcolor{Mycolor2} \textbf{Accuracy}} \\
			\hhline{~|*{4}{-}}
			\multicolumn{1}{c|}{\textbf{Catogram Type}} &
			\textbf{Statistical} &
			\textbf{Geographical} &
			\textbf{Topological} &
			\textbf{Contiguity} \\
			\hline
			Contiguous & \cellcolor{Mycolor3}Variable & \cellcolor{Mycolor3}Variable & Accurate & Yes \\
			\hline
			Non-contiguous & \cellcolor{Mycolor5}\textcolor{white}{Accurate} & \cellcolor{Mycolor5}\textcolor{white}{Shape is accurate} &\cellcolor{Mycolor6}Inaccurate & No   \\
			\hline
			Rectangular & \cellcolor{Mycolor3}Variable & Shape is inaccurate & \cellcolor{Mycolor3}Variable & Yes   \\
			\hline
			Dorling & Accurate & Inaccurate & Inaccurate & No  \\
			\hline
		\end{tabulary}
	}
	\caption{\textcolor{Mycolor3}{\textbf{Trade-off between dimensions}}. \textcolor{Mycolor6}{\textbf{Dimension sacrificed}} in order to improve \textcolor{Mycolor5}{\textbf{target dimension}}'s accuracy.}
	\label{table:accuracy}
\end{table}
}

{
\renewcommand{\arraystretch}{1.5}
\begin{table*}[!htb]
	\centering
	\resizebox{\textwidth}{!}{
		\begin{tabulary}{\textwidth}{|*{4}{l|}r|c|}
			\hhline{~|*{4}{-}|~}
			\multicolumn{1}{c|}{\textbf{Literature}} &
			\textbf{Title} &
			\textbf{Cartogram Type(s)} &
			\textbf{Geographic Region(s)} &
			\textbf{Number of Nodes} &
			\multicolumn{1}{c}{\textbf{Year}} \\
			\hline

@@

Our work extends Demers cartograms. We introduce dynamic topological feature, a river, aiming to improve the interpretation, readability, and geographical accuracy.

>>>>

While we focus on rectangular and Dorling cartograms in this paper, we start with brief definitions of contiguous and non-contiguous cartograms for interested readers: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent neighbors but are also subject to distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity with neighbors to enable expansion or reduction in size while maintaining their polygonal shape \cite{dent2009Cartography}.

@@

{
\renewcommand{\arraystretch}{1.5}
\begin{table}[!tb]
	\centering
	\resizebox{\columnwidth}{!}{
		\begin{tabulary}{\columnwidth}{|*{5}{l|}}
			\hhline{~|*{3}{-}|~}
			\multicolumn{1}{c|}{} & \multicolumn{3}{c|}{\cellcolor{Mycolor2} \textbf{Accuracy}} \\
			\hhline{~|*{4}{-}}
			\multicolumn{1}{c|}{\textbf{Catogram Type}} &
			\textbf{Statistical} &
			\textbf{Geographical} &
			\textbf{Topological} &
			\textbf{Contiguity} \\
			\hline
			Contiguous & \cellcolor{Mycolor3}Variable & \cellcolor{Mycolor3}Variable & Accurate & Yes \\
			\hline
			Non-contiguous & \cellcolor{Mycolor5}\textcolor{white}{Accurate} & \cellcolor{Mycolor5}\textcolor{white}{Shape is accurate} &\cellcolor{Mycolor6}Inaccurate & No   \\
			\hline
			Rectangular & \cellcolor{Mycolor3}Variable & Shape is inaccurate & \cellcolor{Mycolor3}Variable & Yes   \\
			\hline
			Dorling & Accurate & Inaccurate & Inaccurate & No  \\
			\hline
		\end{tabulary}
	}
	\caption{\textcolor{Mycolor3}{\textbf{Trade-off between dimensions}}. \textcolor{Mycolor6}{\textbf{Dimension sacrificed}} in order to improve \textcolor{Mycolor5}{\textbf{target dimension}}'s accuracy.}
	\label{table:accuracy}
\end{table}
}

{
\renewcommand{\arraystretch}{1.5}
\begin{table*}[!tb]
	\centering
	\resizebox{\textwidth}{!}{
		\begin{tabulary}{\textwidth}{|*{4}{l|}r|c|}
			\hhline{~|*{4}{-}|~}
			\multicolumn{1}{c|}{\textbf{Literature}} &
			\textbf{Title} &
			\textbf{Cartogram Type(s)} &
			\textbf{Geographic Region(s)} &
			\textbf{Number of Nodes} &
			\multicolumn{1}{c}{\textbf{Year}} \\
			\hline

@@

One of the disadvantages of Dorling and Demers cartograms is the legibility. The layout algorithms may displace regions far from their original position and making them more difficult to interpret. Our work extends Demers cartograms. We introduce a new type of dynamic topological feature, a river, aiming to improve the interpretation, readability, and geographical accuracy of this class of cartograms.
